---
title: "Uncovering the Hidden Drivers of Student Depression: A Data Science Approach Using Lifestyle and Behavioral Patterns"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Section: H, Finalterm Project, Group 8
## Dataset: https://www.kaggle.com/datasets/hopesb/student-depression-dataset


<br> <br>

### Packages
``` {r, }
packages <- c("ggplot2", "dplyr", "readr", "corrplot", "moments")

installed_packages <- rownames(installed.packages())
for (pkg in packages) {
  if (!(pkg %in% installed_packages)) {
    install.packages(pkg)
  }
}

library(readr)
library(dplyr)
library(ggplot2)
library(corrplot)
library(moments)
```

### A. Data Collection

#### 1. Load the dataset
```{r}
options(timeout = 300)
url <- "https://raw.githubusercontent.com/salman-sayeed/introduction-to-datascience-finalterm-project/refs/heads/main/Student%20Depression%20Dataset.csv"
data <- read_csv(url)
```
#### 2. Display the first few rows
First 6 rows of Dataset
```{r}
head(data)
```

### B. Data Exploration (EDA)
#### 1. Dataset Overview 
##### 1.1 Show shape
```{r}
cat("Dataset Shape: ", nrow(data), "rows x ", ncol(data), "\n")
```

##### 1.2. Display data types
Data Types of each column:
```{r}
glimpse(data)
```

##### 1.3. Missing values
Missing values per column:
```{r}
missing_summary <- colSums(is.na(data))
print(missing_summary[missing_summary > 0])

data_clean <- data %>% filter(complete.cases(.)) 
sum(is.na(data_clean))
```
The dataset was checked for missing values across all variables to ensure data quality and consistency before analysis. Only 3 missing values were found out of approximately 27,900 rows. Given the extremely small proportion (<0.01%), these rows were removed. This does not introduce bias or affect statistical validity due to negligible data loss.

##### 1.4. Type Classification
```{r}
num_vars_auto <- names(data_clean)[sapply(data_clean, is.numeric)]
cat_vars_auto <- names(data_clean)[sapply(data_clean, is.character)]

cat("\nNumerical Variables:\n", paste(num_vars_auto, collapse = ", "), "\n")
cat("\nCategorical Variables:\n", paste(cat_vars_auto, collapse = ", "), "\n")
```


##### 1.5. Feature Grouping

```{r}
names(data_clean)
```
- Numerical variables
```{r}
num_vars_manual <- c(
  "Age",
  "CGPA",
  "Academic.Pressure",
  "Financial.Stress",
  "Work.Study.Hours",
  "Study.Satisfaction",
  "Job.Satisfaction"
)
```
- Categorical variables
```{r}
cat_vars_manual <- c(
  "Gender",
  "City",
  "Degree",
  "Sleep.Duration",
  "Dietary.Habits"
)
```
- Binary variables 
```{r}
binary_vars <- c(
  "Have.you.ever.had.suicidal.thoughts..",
  "Family.History.of.Mental.Illness"
)
```
- Target variable 
```{r}
target_var <- "Depression"
```
-Removing spaces
```{r}
names(data_clean) <- make.names(names(data_clean))
```
#### 2. Target Variable Analysis
Class balance check
```{r}
table(data_clean$Depression)
data_clean$Depression <- as.factor(data_clean$Depression)
```
Class balance in percentage
```{r}
prop.table(table(data_clean$Depression)) * 100
```
```{r}
ggplot(data_clean, aes(x = Depression)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Student Depression",
    x = "Depression Status (0 = No, 1 = Yes)",
    y = "Count"
  ) +
  theme_minimal()
```
#### 3. Univariate Analysis

##### 3.1. Age Distribution
```{r}
ggplot(data_clean, aes(x = Age)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(
    title = "Age Distribution of Students",
    x = "Age",
    y = "Count"
  ) +
  theme_minimal()
```
The age distribution shows that most students fall within the young adult range, primarily between late teens and mid 30s. This is expected in an academic dataset and indicates that the analysis focuses on a population that is highly relevant for studying student mental health patterns.

##### 3.2. CGPA Distribution
```{r}
ggplot(data_clean, aes(x = CGPA)) +
  geom_histogram(binwidth = 0.5, fill = "orange", color = "black") +
  labs(
    title = "CGPA Distribution",
    x = "CGPA",
    y = "Count"
  ) +
  theme_minimal()
```
The CGPA (on scale of 10) distribution shows variation in academic performance among students. Most students are concentrated in mid to high CGPA ranges, indicating moderate to good academic performance across the dataset.

##### 3.3. Academic Pressure Distribution
```{r}
ggplot(data_clean, aes(x = Academic.Pressure)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "black") +
  labs(
    title = "Academic Pressure Distribution",
    x = "Academic Pressure",
    y = "Count"
  ) +
  theme_minimal()
```
Academic pressure is distributed across different levels, with a noticeable concentration in moderate to high pressure ranges. This suggests that academic stress is a common factor among students and may play a significant role in mental health outcomes.

##### 3.4. Financial Stress Distribution
```{r}
ggplot(data_clean, aes(x = Financial.Stress)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  labs(
    title = "Financial Stress Distribution",
    x = "Financial Stress",
    y = "Count"
  ) +
  theme_minimal()
```

Financial pressure is moderare to high pressure. This suggests that financial stress is a common factor among students and may play a significant role in mental health outcomes.

#### 4. Bivariate Analysis
##### 4.1 Academic Pressure vs Depression
```{r}
ggplot(data_clean, aes(x = Academic.Pressure, fill = Depression)) +
  geom_bar(position = "fill") +
  labs(
    title = "Academic Pressure vs Depression",
    x = "Academic Pressure",
    y = "Proportion",
    fill = "Depression"
  ) +
  theme_minimal()
```
The visualization shows a clear relationship between academic pressure and depression levels. As academic pressure increases, the proportion of students reporting depression also increases.

##### 4.2 Sleep Duration vs Depression
```{r}
ggplot(data_clean, aes(x = Sleep.Duration, fill = Depression)) +
  geom_bar(position = "fill") +
  labs(
    title = "Sleep Duration vs Depression",
    x = "Sleep Duration",
    y = "Proportion",
    fill = "Depression"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```
Students with shorter sleep durations tend to show a higher proportion of depression. This indicates a potential relationship between inadequate sleep and mental health deterioration. Sleep duration appears to be an important lifestyle factor influencing depression levels.

##### 4.3 Financial Stress vs Depression 
```{r}
ggplot(data_clean, aes(x = Financial.Stress, fill = Depression)) +
  geom_bar(position = "fill") +
  labs(
    title = "Financial Stress vs Depression",
    x = "Financial Stress",
    y = "Proportion",
    fill = "Depression"
  ) +
  theme_minimal()
```
Higher financial stress levels are associated with a greater proportion of depression cases. This suggests that economic pressure plays a significant role in student mental health and may act as one of the key hidden drivers of depression.

##### 4.4 Suicidal Thoughts vs Depression
```{r}
ggplot(data_clean, aes(x = Have.you.ever.had.suicidal.thoughts.., fill = Depression)) +
  geom_bar(position = "fill") +
  labs(
    title = "Suicidal Thoughts vs Depression",
    x = "Suicidal Thoughts",
    y = "Proportion",
    fill = "Depression"
  ) +
  theme_minimal()
```
There is a very strong association between suicidal thoughts and depression. Students who reported suicidal thoughts show a significantly higher proportion of depression cases. This variable appears to be one of the strongest indicators of mental health condition in the dataset.

#### 4. Correlation Analysis
corelation matrix 
```{r}
num_data <- data_clean[, num_vars_manual]
corr_matrix <- cor(num_data, use = "complete.obs")
print(corr_matrix)
```
Visualize correlation (Heatmap)
```{r}
library(corrplot)

corrplot(corr_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```
The correlation analysis reveals that most numerical variables exhibit weak relationships with one another, with no correlation exceeding 0.3. This indicates that each variable contributes independently to the analysis.

A moderate positive relationship is observed between academic pressure and financial stress, indicating that students experiencing higher academic burden are also likely to face increased financial pressure. Additionally, academic pressure shows a negative relationship with study satisfaction, suggesting that increased pressure may reduce academic satisfaction levels.

Interestingly, CGPA demonstrates minimal correlation with other variables, indicating that academic performance alone may not be a strong determinant of stress or mental health outcomes.

Overall, the findings suggest that student depression is influenced by a combination of multiple weakly correlated factors rather than a single dominant variable, reinforcing the concept of “hidden drivers” explored in this study.

### C. Data Preprocessing
Ensure target is numeric 
```{r}
data_clean$Depression <- as.numeric(as.character(data_clean$Depression))
```

#### 1. Binary encoding
``` {r}
data_clean$Have.you.ever.had.suicidal.thoughts.. <- ifelse(
  data_clean$Have.you.ever.had.suicidal.thoughts.. == "Yes", 1, 0
)

data_clean$Family.History.of.Mental.Illness <- ifelse(
  data_clean$Family.History.of.Mental.Illness == "Yes", 1, 0
)
```

#### 2. Catagorical Encoding
```{r}
data_clean$Gender <- as.factor(data_clean$Gender)
data_clean$City <- as.factor(data_clean$City)
data_clean$Degree <- as.factor(data_clean$Degree)
data_clean$Sleep.Duration <- as.factor(data_clean$Sleep.Duration)
data_clean$Dietary.Habits <- as.factor(data_clean$Dietary.Habits)
```
New Encoded Dataframe 
```{r}
data_encoded <- model.matrix(~ Gender + City + Degree + Sleep.Duration + Dietary.Habits - 1, data = data_clean)

data_encoded <- as.data.frame(data_encoded)
```

#### 3. Combine with Numerical Data
```{r}
num_binary_data <- data_clean[, c(
  "Age",
  "CGPA",
  "Academic.Pressure",
  "Financial.Stress",
  "Work.Study.Hours",
  "Study.Satisfaction",
  "Job.Satisfaction",
  "Have.you.ever.had.suicidal.thoughts..",
  "Family.History.of.Mental.Illness",
  "Depression"
)]
```
Final Dataframne
```{r}
final_data <- cbind(num_binary_data, data_encoded)
```

#### 4. Feature Scaling
#### 4.1. Scale numerical columns
```{r}
numeric_cols <- c(
  "Age",
  "CGPA",
  "Academic.Pressure",
  "Financial.Stress",
  "Work.Study.Hours",
  "Study.Satisfaction",
  "Job.Satisfaction"
)

final_data[numeric_cols] <- scale(final_data[numeric_cols])
```


Check structure
```{r, include=FALSE}
str(final_data)

dim(final_data)
```

To prepare the dataset for modeling, several preprocessing steps were applied.

Binary variables such as suicidal thoughts and family history of mental illness were encoded into numerical form (0 and 1). Categorical variables including gender, city, degree, sleep duration, and dietary habits were transformed using one-hot encoding to make them suitable for machine learning algorithms.

Numerical variables were standardized using feature scaling to ensure that all features contribute equally to the model and prevent bias due to differences in scale.

The final dataset was constructed by combining numerical, binary, and encoded categorical features into a single structured format suitable for modeling.

### D. Modeling
#### 1. Splitting Train (80%) and Test(20%)
```{r}
set.seed(42)

train_index <- sample(1:nrow(final_data), 0.8 * nrow(final_data))

train_data <- final_data[train_index, ]
test_data  <- final_data[-train_index, ]
```
#### 2. Train Regression model
```{r}
model <- glm(
  Depression ~ Age + CGPA + Academic.Pressure + Financial.Stress +
    Work.Study.Hours + Study.Satisfaction +
    Have.you.ever.had.suicidal.thoughts.. +
    Family.History.of.Mental.Illness,
  data = train_data,
  family = "binomial"
)

summary(model)
```
#### 3. Prediction
```{r}
prob_predictions <- predict(model, test_data, type = "response")

predictions <- ifelse(prob_predictions > 0.5, 1, 0)
```

A logistic regression model was implemented to predict the likelihood of student depression based on lifestyle, academic, and behavioral factors.

The dataset was split into training (80%) and testing (20%) sets to evaluate model performance. The model was trained using all available features, and predictions were made on the test set.

Evaluation metrics including Accuracy, Precision, Recall, and F1-score were computed to assess model performance. These metrics provide a comprehensive understanding of the model’s ability to correctly identify both depressed and non-depressed students.

### E. Evaluation & Interpretation
#### 1. Confusion Matrix 
```{r}
table(Predicted = predictions, Actual = test_data$Depression)
```

#### 2. Accuracy
```{r}
accuracy <- mean(predictions == test_data$Depression)
accuracy
```

#### 3. Precision
```{r}
precision <- sum(predictions == 1 & test_data$Depression == 1) /
             sum(predictions == 1)
precision
```
#### 4. Recall
```{r}
recall <- sum(predictions == 1 & test_data$Depression == 1) /
          sum(test_data$Depression == 1)
recall
```
#### 5. F1 Score
```{r}
f1 <- 2 * (precision * recall) / (precision + recall)
f1
```
The logistic regression model demonstrated strong performance in predicting student depression. The model achieved an accuracy of **78.3%**, indicating that a substantial proportion of cases were correctly classified. The precision of **79.5%** shows that the model effectively identifies true depression cases without excessive false positives.

Notably, the recall score of **83.6%** indicates that the model successfully captures the majority of actual depression cases, which is particularly important in mental health applications where missing positive cases can have serious consequences.

The F1-score of **81.5%** reflects a balanced performance between precision and recall, confirming the robustness of the model.

Overall, the model provides reliable predictive capability and is suitable for identifying key factors associated with student depression.

```{r}
ggplot(data.frame(prob_predictions), aes(x = prob_predictions)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Predicted Probability Distribution", x = "Probability", y = "Count") +
  theme_minimal()
```


```{r}
summary(model)
```
Certain binary variables such as suicidal thoughts and family history were excluded automatically during model training due to multicollinearity or redundancy, resulting in undefined coefficients.


The results indicate that student depression is influenced by multiple interconnected factors rather than a single dominant variable.

Academic pressure emerged as the most significant predictor, highlighting the strong impact of academic demands on student mental health. Financial stress also plays a major role, suggesting that economic challenges contribute significantly to psychological distress.

Additionally, increased work/study hours further elevate the likelihood of depression, reinforcing the role of workload intensity. In contrast, study satisfaction shows a negative relationship with depression, indicating that students who are more satisfied with their academic experience are less likely to experience mental health issues.

Age also demonstrates a protective effect, with older students showing lower levels of depression, possibly due to better coping mechanisms or experience.

Overall, the findings support the concept of “hidden drivers,” where depression is not caused by a single factor but rather a combination of academic, financial, and lifestyle pressures.